---
title: "EDUC 641 Syllabus"
site: distill::distill_website
output:
  html_document:
    toc: TRUE
    toc_depth: 3
    toc_float: TRUE
    theme: cosmo
    css: educquant.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Course Overview

This course provides a first introduction to quantitative data analysis in education and the social sciences. It is part of a three course sequence intended to provide a toolkit of statistical concepts, methods and their implementation to producers of applied research in education and other social sciences. This course is intended for those who will pursue additional statistics courses with the aim of ultimately conducting their own, independent research. The course is organized around the principle that research design depends in part on researchers' substantive questions and their quantitative data available to answer these question. In this introductory course, we will focus on describing categorical and continuous data and quantifying the relationship between categorical and continuous data. Students will form a solid foundation for frequentist, inferential statistics (and some of critiques of this model). The course seeks to blend a conceptual, mathematical and applied understanding of basic statistical concepts. At the core of our pedagogical approach is the belief that students learn statistical analysis by **doing** statistical analysis. *This course (or substitute) is a pre-requisite for EDUC 643*.

# Meeting Time and Location

# Instructors

# Student Learning Outcomes
By the end of this term, it is expected that students will be able to:

1. Articulate the principles of responsible and ethical quantitative research
2. Describe, summarize and visualize quantitative data that are categorical and continuous
3. Describe, summarize and visualize the relationships between quantitative data that are categorical and continuous
4. Conduct a frequentist null-hypothesis significance test of the relationships between simple categorical and continuous data
5. Describe strategies to improve the replicability and generalizability of quantitative research
6. Use th R programming language to conduct all such analyses


# Textbooks and Reading Materials

## Textbook
We will primarily be referring to chapters in [Learning Statistics with R](https://learningstatisticswithr-bookdown.netlify.com/index.html) (**LSWR**) by Danielle Navarro. This textbook is available for free online. You may choose to purchase a paper copy if you wish, but it is not required. 

There are some additional readings assignments on the troubling historical origins of modern statistics that appear in the detailed [schedule](./schedule.html). 

### Other text resources
There are literally thousands of high-quality introductory statistics textbooks. We have particularly found...

## R and RStudio

Students must have the latest version of R, which can be downloaded [here](https://ftp.osuosl.org/pub/cran/). It is strongly recommended that students also download the RStudio GUI, available [here](https://www.rstudio.com/products/rstudio/download/#download). Both softwares are free. We will provide tutorials on R/RStudio installation and they are also accessible [here](.slides/tutorial.html).

### Resources for R and RStudio

While we will teach you how to effectively use R and RStudio to conduct analyses, one of the key skills required to use R is the ability to find answers on your own. Many common questions or problems are either available on blogs or have been asked and answered in discussion forums already. Finding and deciphering those answers is an important skill you should seek to hone. *You will never remember all of the programming commands!*

Here are some sites where you can find the answers to many R questions and learn new tricks:

 - [_YaRrr, the Pirate's Guide to R_](https://bookdown.org/ndphillips/YaRrr/) by Nathaniel Phillips
 - [_R for Data Science_](https://r4ds.had.co.nz/) by Hadley Wickham
 - [_R Cookbook_](http://www.cookbook-r.com/) by Winston Chang
 - [_An Introduction to Statistical Learning_](http://www-bcf.usc.edu/~gareth/ISL) by Gareth James, Daniela Witten, Trevor Hastie and Robert Tibshirani
 - [Cheat Sheets](https://www.rstudio.com/resources/cheatsheets) 
 - [Quick-R](https://www.statmethods.net) 
 - [StackOverflow](https://stackoverflow.com/questions/tagged/r)
 - [Stack Exchange](https://stats.stackexchange.com)


# Schedule

For more details, see [here](./schedule.html). 

```{r, echo=FALSE, results='asis'}
d <- readxl::read_xlsx(here::here("./data", "syllabus.xlsx"))
d %>% 
  knitr::kable()
```


# Grading Components and Criteria 
Final grades will be based on the following components:

- Quizzes: X%
- Assignments: Y%
- Final: Z%

## Quizzes
Five classes will begin with a very short multiple-choice quiz that is designed to test your knowledge of the theoretical principles underlying the statistics we are studying for the week. While some may feel that this is overly paternalistic, research evidence shows that frequent quizzing increases learning (see a summary of [one study](https://www.nytimes.com/2013/11/21/education/frequent-tests-can-enhance-college-learning-study-finds.html) from the University of Texas). Quizzes will be open book and notes (but not laptop/computer).

## Assignments


## Final

# Student Engagement Inventory
Graduate students are expected to perform work of high quality and quantity, typically with forty hours of student engagement for each student credit hour. For this course, the following table shows the number of hours a typical student would expect to spend in each of the following activities:

| Educational activity     |     Hours     |    Explanatory comments
|--------------------------------------------------------------------------
| Class attendance         | 30            | TBD frequency
| Class reading and prep   | 20            | Includes reading and review of slides
| Homework Assignments     | 40            | Homework assignments will take 10 hours each (on avg.)
| Final                    | 30            | Includes familiarization with data, data analysis, preparation of displays and writing
| **Total hours**          | 120           | **These are approximations.** Reading and *especially* analytic time will vary per individual


## Role of the GEs  
## Attendance and Absence Guidelines  
## Expected Classroom Behavior
## Diversity, Equity, and Inclusion  
## Indigenous Recognition Statement  
## Using Pronouns and Personal Preference  
## Your Well-being
## Reporting Obligations  
## More per university requirements...



