---
title: "EDUC 641 Continuous Data"
author: "Professor Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
  html_document:
    keep_md: true
---
class: inverse, center, middle

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.align = "center",
                      fig.height = 3)

if (!require(pacman)) install.packages('pacman', repos = 'https://cran.rstudio.com')
pacman::p_load(tidyverse, here, knitr, dplyr, ggplot2, blogdown, rio)
```

```{r, echo = F}
who <- import(here::here("data", "life_expectancy.csv")) %>% 
  janitor::clean_names() %>% 
  # filter(year == 2015) %>% 
  select(country, status, year, life_expectancy) %>% 
  rename(region = country) %>% 
  mutate(life_expectancy = round(life_expectancy, digits = 0))

world <- map_data("world")

who1<- unique(who$region)
world1 <- unique(world$region)


ggplot(subset(who, year == 2013), aes(x = life_expectancy)) +
  geom_density()
```

# Agenda
* Sampling Distributions
* Central Limit Theorem

---
# Review
## Distributions

Characterized by:

* Central Tendency
+ Mean
+ Median
+ Mode

* Variability
+ Range
+ Interquartile Range
+ Variance
+ Standard Deviation

* Shape
+ Skew
+ Kurtosis

---
# Review
## Normal Distributions

A normal distribution is a **theoretical** distribution with the following characteristics:

Central Tendency:
+ The mean, median, and mode are equivalent values.

Variability:
+ 68% of observations are within 1 SD
+ 95% of  observations are within 2 SDs
+ 99% of observations are within 3 SDs

Shape: 
+ Symmetrical

---
# Population vs. Sample Distributions

Many phenomena in the natural and social sciences resemble a normal distribution. 

Particularly as distributions get smaller, they can be idiosyncratic or have

```{r}




```

# Life Expectancy Data
Suppose you are working for the World Health Organization and are investigating life expectancy across different regions. 

> Using this dataset, we can ask questions like:
> * How does life expectancy compare in developing vs. developed regions?
> * What is the relationship between average years of schooling and life expectancy?

---
#