---
title: "EDUC 641 Continuous Data"
author: "Professor Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  xaringan::moon_reader:
    css: ['default','new.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
  html_document:
    keep_md: true
---
class: inverse, center, middle

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.align = "center",
                      fig.height = 3)

if (!require(pacman)) install.packages('pacman', repos = 'https://cran.rstudio.com')
pacman::p_load(tidyverse, here, knitr, dplyr, ggplot2, blogdown, rio)
```

```{r, echo = F}
who <- import(here::here("data", "life_expectancy.csv")) %>% 
  janitor::clean_names() %>% 
  filter(year == 2015) %>% 
  select(country, status, life_expectancy) %>% 
  rename(region = country) %>% 
  mutate(life_expectancy = round(life_expectancy, digits = 0))

world <- map_data("world")

who1<- unique(who$region)
world1 <- unique(world$region)

 
 getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

```

# Agenda
* Objectives
* Life Expectancy Data
* Measures of Central Tendency
* Measures of Variability

---
# Life Expectancy Data
Suppose you are working for the World Health Organization and are investigating life expectancy across different regions. 

> Using this dataset, we can ask questions like:
> * How does life expectancy compare in developing vs. developed regions?
> * Do individuals living in countries with more total years of attendance in school experience, on average, higher life expectancy??


**But**, before making comparisons between social/economic conditions, we want to start with describing the full dataset, looking at all 183 nations in the dataset.

**Our Task:** Describe the distribution of regional life expectancy in 2015.

---
# What is a distribution?

A distribution describes the number of observations of variable that take a particular value.

```{r, echo = F, fig.height = 6, fig.width = 10}
d <- who %>% 
      filter(str_length(region) <= 12) %>% 
      sample_n(60)

d_hist <- hist(d$life_expectancy, plot = FALSE)

d$x_bin <- cut(d$life_expectancy, 
               breaks = d_hist$breaks,
               label = FALSE)

d$x_mid <- d_hist$mids[d$x_bin]
d$y <- 1

d <- d %>% 
  group_by(x_bin) %>% 
  mutate(y = cumsum(y))

ggplot() +
  geom_label(data = d, 
             aes(x = x_mid,
                 y = y,
                 label = region),
             size = 4) +
  theme_bw(base_size = 12) +
  theme(panel.grid = element_blank()) +
  labs(x = "Average Life Expectancy", y = "Count")

```

---
# Describing a Distribution

There are many ways to describe the characteristics of a distribution. 

Today we are reviewing:

* Measures of Central Tendency
  + Mean
  + Median
  + Mode
  
* Spread or Variability
  + Variance
  + Standard Deviation
  + Interquartile Range
  + Range

* Measures of Skewness
* Measures of Kurtosis
  
---
# But first, summation notation

Summation (or sigma) notation is used to provide a concise expression for the sum of observations.

$$\sum_{i=i}^{n} x_{i}$$
---
# Central Tendency
## Mean
* Represents the **average**, or the sum of all observations divided by the number of observations.
* One of the most common forms of central tendency.
* Represented by the Greek symbol $\mu$.

> $$\bar{X} = \frac{1}{n}{\sum_{i=i}^{n} X_{i}}$$
 
> $$mean(71, 73, 76, 78, 79) = \frac{71 + 73 + 76 + 78 + 79}{5} = 75.4$$

The mean national life expectancy in 2015 is **71.64** years.
---
# Central Tendency
## Median
* Represents the **midpoint** of the distribution.
* Also called the *50th percentile* of a distribution or *2nd quartile*, meaning half of the observations are above the median and half are below.
* Particularly useful for describing the central tendency of skewed distributions (more on this later).
* To find the median, arrange the scores in ascending order and identify the middle value. If there are two middle values (even number of observations), use the average of the two.

> $$median(71, 73, 76, 78, 79) = 76$$
> $$median(71, 73, 76, 77, 78, 79) = 76.5$$
```{r}
median(who$life_expectancy)
```

The median national life expectancy in 2015 was **74** years.

---
# Central Tendency
## Mode
The mode simply refers to the **most frequent value** in the data.

What is the mode of the following data?

> 78, 79, 78, 77, 80, 79, 76, 74, 79, 78, 77, 78, 80, 79, 78, 76, 79, 77, 78, 78

It's easier if we sort the data first.

> 74, 76, 76, 77, 77, 77, **78, 78, 78, 78, 78, 78, 78,** 79, 79, 79, 79, 79, 80, 80

78 is the most frequent value in the data. Therefore, the mode is 78.

---
# Central Tendency
## Mode

Sometimes there may be more than one mode. Consider the following example:

> 4, 6, 6, 7, 7, 7, **8, 8, 8, 8, 8, 8, 8, 9, 9, 9, 9, 9, 9, 9,** 10, 10

Both 8 and 9 are observed seven times. In this case, the data can be considered **bimodal** and both modes would be reported.

In general, the mode is rarely used and is particularly vulnerable to idiosyncratic patterns in the data.

---

# Response Opportunity
## Comparing Distributions

These distributions have the same central tendency $(\mu = 100)$.

How are they different?

```{r, echo = F}
narrow <- tibble(values = rnorm(10000, mean = 100, sd = 1)) %>% 
  ggplot(aes(values)) +
  geom_density(adjust = 3) +
  theme_classic() +
  geom_vline(xintercept = 100, color = "red")

wide <- tibble(values = rnorm(10000, mean = 100, sd = 10)) %>% 
  ggplot(aes(values)) +
  geom_density(adjust = 3) +
  theme_classic() +
  geom_vline(xintercept = 100, color = "red")

gridExtra::grid.arrange(narrow, wide, nrow = 1)
```

---
# Variability

* Central tendency is only one component of a distribution.

* What if we want to know how much variation there is in life expectancy across nations?
  + How much does typical observation deviate from the mean?

* **Measures of Variability:**
  + Range
  + Interquartile Range
  + Variance
  + Standard Deviation

---
# Range

* Represents the difference between the highest value and lowest value in the dataset.
* Can provide a rough estimate of spread

> $$range(71, 73, 76, 78, 79) = 79 - 71 = 8$$

* Very vulnerable to outliers. One observation can make a big difference!

> $$range(55, 73, 76, 78, 79) = 79 - 55 = 24$$

---
# Interquartile Range (IQR)
* Represents the difference between the 1st quartile (25th percentile) and the 3rd quartile (75th percentile).

* **Percentile** refers to the percentage of observations that fall at or below that score.
  + e.g., 25% of observed national life expectancies fall at or below 66 years of age.

* IQR summarizes the range of "average" scores in the dataset.

```{r, echo = F}
ggplot(who, aes(x = life_expectancy)) +
  geom_density() +
  geom_vline(xintercept = quantile(who$life_expectancy, .25)) +
  geom_label(
    x = quantile(who$life_expectancy, .25),
    y = .03,
    label = paste("25th %ile =", quantile(who$life_expectancy, .25))
  ) +
  geom_vline(xintercept = quantile(who$life_expectancy, .75)) +
  geom_label(
    x = quantile(who$life_expectancy, .75),
    y = .03,
    label = paste("75th %ile =", quantile(who$life_expectancy, .75))
  ) +
  geom_segment(
    x = quantile(who$life_expectancy, .25),
    xend = quantile(who$life_expectancy, .75),
    y = .015,
    yend = .015,
    linetype = 2
  ) +
  geom_label(x = 71.5,
             y = 0.01,
             label = "IQR = 11") +
  theme_classic()
  
```

---
# Variance
* Represents the **average squared deviation** of each observation from the mean.

$$s^{2} = \frac{\sum_{i=i}^{n} x_{i}}{n}$$

* Why do we square the deviation?

---
# Variance
Here are some national life expectancies centered around the mean. Their value represents their difference relative to the sample mean.
  
```{r, echo = F, fig.height = 5, fig.width = 8}
who$life_cnt <- round((who$life_expectancy - mean(who$life_expectancy)))
who$life_type <- ifelse(who$life_cnt < 0, "below", "above")
who$region <- as.factor(who$region)

set.seed(154)
var_plot <- who %>%
  sample_frac(.10) %>% 
  arrange(life_cnt) %>% 
  ggplot(aes(x=reorder(region, desc(life_cnt)), y=life_cnt, label=life_cnt)) +
geom_bar(stat='identity', aes(fill=life_type), width=.5) +
scale_fill_manual(name="Life Expectancy (deviation)",
labels = c("Above Average", "Below Average"),
values = c("above"="#00ba38", "below"="#0b8fd3")) +
labs(subtitle="Mean-centered life expectancy for random subset of nations") +
  coord_flip() +
  theme_minimal()

var_plot
```
Some observations fall below the mean so their deviation is a negative value. If we took the average of our positive and negative deviations, we'd just get zero!

---
# Variance
Since any number squared is positive, let's try squaring our deviations.

```{r, echo = F, fig.height = 5, fig.width = 8}
who$life_cnt_sq <- round((who$life_expectancy - mean(who$life_expectancy))^2)

set.seed(154)
var_plot_sq <- who %>% 
  sample_frac(.10) %>% 
  arrange(life_cnt) %>% 
  ggplot(aes(x=reorder(region, desc(life_cnt)), y=life_cnt_sq, label=life_cnt)) +
geom_bar(stat='identity', aes(fill=life_type), width=.5) +
scale_fill_manual(name="Life Expectancy (deviation)",
labels = c("Above Average", "Below Average"),
values = c("above"="#00ba38", "below"="#0b8fd3")) +
labs(subtitle="Mean-centered life expectancy (squared) for random subset of nations") +
  theme_minimal() +
  coord_flip()

var_plot_sq
```
Now we can take the average of our squared deviations and get a non-zero value!
---
# Variance

Our variance (or average squared deviation) is approximately 66.43.

```{r, echo = F, fig.height = 5, fig.width = 8}
var_plot_sq +
  geom_hline(yintercept = mean(who$life_cnt_sq), color = "red")
```

By itself, a variance of 66.43 is not very meaningful to us. It does not reflect the raw units of our original scale (it represents squared units).
---
# Standard Deviation

* If we take the square root of our sample variance, we can summarize average deviations back on our raw scale. 

* The standard deviation represents the **positive square root of the variance**.
$$ s = \sqrt{s^2} $$
$$ \sqrt{66.43} = 8.15 $$

The standard deviation of life expectancy in 2015 was approximately 8.15 *years*.

---
# Standard Deviation

Here is our original plot with our <span style="color: red;">standard deviation</span> plotted below and above the mean (i.e., -1 SD, +1 SD).

```{r, echo = F, fig.height = 5, fig.width = 8}
var_plot +
  geom_hline(yintercept = 8.15, color = "red", linetype = 2) +
  geom_hline(yintercept = -8.15, color = "red", linetype = 2)
```

The standard deviation of life expectancy in 2015 was approximately 8.15 *years*.

---
# Z-unit Distribution

* A distribution can have any mean and any (positive) standard deviation.
* A z-distribution is a useful distribution with a mean of 0 $(\mu)$ and a standard deviation $(\sigma)$ of 1.

```{r echo=FALSE}
ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  stat_function(fun = dnorm) +
  scale_x_continuous(limits = c(-4, 4),
                     breaks = c(seq(-4, 4, 1))) +
  ylim(0, 0.4) +
  theme_classic() +
  ggtitle("Z Distribution", subtitle =  "m = 0, sd = 1") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

---
# Transforming Distributions
* Any distribution can be transformed into a z distribution using a linear function.

Each observation is transformed into a **z-score** using the following formula:
$$z = \frac{X - \mu}{\sigma}$$

* A z-score is calculated by **subtracting the mean** from each value and **dividing by the standard deviation**.

* An observation's z-score value is equal to its distance from the mean, in standard deviation units.

---
# Transforming Distributions

Here is a histogram of our life expectancy data.
```{r, echo = F}
hist(who$life_expectancy)
```

We are going to create a new variable called `life_expectancy_zscore` using the linear function described on the previous slide.
```{r}
who$life_expectancy_zscore <- (who$life_expectancy - mean(who$life_expectancy))/sd(who$life_expectancy)
```

---
# Transforming Distributions

Let's look at the first few transformed values.
```{r}
head(who$life_expectancy_zscore)
```

```{r}
## Histogram of the new z-scores
hist(who$life_expectancy_zscore)
```

Our new mean is 0.
Our new standard deviation is 1.

---
# Distributional Shape
## Skewness
* Not all distributions in the real world are symmetrical. Often they have asymmetry or **skewness**.
* Skewness describes how much a distribution is "bunched up" to the right or left.

```{r, echo = F}
set.seed(1000)
 N <- 10000
 x <- tibble(pos_skew = rbeta(10000, 1.5 ,5),
                 neg_skew = rbeta(10000, 5, 1.5),
                 central = rbeta(10000, 5 , 5)) %>% 
     mutate(across(everything()), round(., digits = 2))
 
pos_skew <- ggplot(x, aes(pos_skew)) +
  geom_density(adjust = 4) +
  xlim(0, 1) + 
  ylim(0, 2.5) +
  ggtitle("Positive Skew") +
  theme_classic() +
  theme(axis.title = element_blank())

neg_skew <- ggplot(x, aes(neg_skew)) +
  geom_density(adjust = 4) +
  xlim(0, 1) + 
  ylim(0, 2.5) +
  ggtitle("Negative Skew") +
  theme_classic() +
  theme(axis.title = element_blank())

normal <- ggplot(x, aes(central)) +
  geom_density(adjust = 2) +
  xlim(0, 1) + 
  ylim(0, 2.5) +
  ggtitle("No Skew") +
  theme_classic() +
  theme(axis.title = element_blank())

gridExtra::grid.arrange(pos_skew, normal, neg_skew, nrow = 1)
```

* Positive Skew - Distribution is skewed to the left with a "positive-pointing finger."
* Negative Skew - Distribution is skewed to the right with a "negative-pointing finger."

---
# Distributional Shape
## Kurtosis

* Kurtosis describes how much the values concentrate around the center.
 + Leptokurtic distributions have more values around the center and have a "taller" peak.
 + Platykurtic distributions do not have a prominent peak and have a flatter top.
 
```{r, echo = F}
# Fleischmann coeffiecients to generate a distribution with kurtosis = 4
a <- 0
b <- 0.7373813
c <- 0
d <- 0.08092509

# Fleischmann coeffiecients to generate a distribution with kurtosis = -1
e <- 0
f <- 1.221009
g <- 0
h <- -0.08015816

k <- tibble(central = rnorm(10000, 0, 1)) %>% 
  mutate(lepto = a + b*central + c*(central^2) + d*(central^3),
         platy = e + f*central + g*(central^2) + h*(central^3))
         

platy <- ggplot(k, aes(platy)) +
  geom_density(adjust = 4) +
  xlim(-3, 3) +
  ylim(0, .5) +
  theme_classic() +
  theme(axis.title = element_blank()) +
  ggtitle("Platykurtic")

lepto <- ggplot(k, aes(lepto)) +
  geom_density(adjust = 4) +
  xlim(-3, 3) +
  ylim(0, .5) +
  theme_classic() +
  theme(axis.title = element_blank()) +
  ggtitle("Leptokurtic")

central <- ggplot(k, aes(central)) +
  geom_density(adjust = 4) +
  xlim(-3, 3) +
  ylim(0, .5) +
  theme_classic() +
  theme(axis.title = element_blank()) +
  ggtitle("Mesokurtic (No Kurtosis)")


gridExtra::grid.arrange(platy, central, lepto, nrow = 1)

```
---
# Evaluating Skewness and Kurtosis

* Skewness and kurtosis can be evaluated visually and summarized quantitatively.

  + Positive skew (skewed to the left) ----> Skewness > 0
  + Negative skew (skewed to the right) ----> Skewness > 0

  + Leptokurtic ----> Kurtosis > 0
  + Platykurtic ----> Kurtosis < 0
  
---
# Summarizing Data
## Applied Example

* Here is a density plot of the WHO life expectancy data for 2015. 
* How would you characterize the distribution?

```{r, echo = F}
ggplot(who, aes(life_expectancy)) +
  geom_density() +
  theme_minimal()
```

* Is there severe skew?
* Is there severe kurtosis?

---
# Summarizing Data
## Applied Example

The distribution of life expectancy has a slightly negative skew.
```{r, echo = F}
ggplot(who, aes(life_expectancy)) +
  geom_density() +
  geom_vline(xintercept = mean(who$life_expectancy), color = "red") +
  geom_vline(xintercept = median(who$life_expectancy), color = "blue") +
  theme_minimal()
```

```{r}
psych::describe(who$life_expectancy) %>% 
  select(n, mean)
```

The <span style="color: red;">mean</span> national life expectancy in 2015 was **71.62** years.

The <span style="color: blue;">median</span> national life expectancy in 2015 was **74** years.

* The median appears to better capture the "center" of our distribution.

---


### Applied Coding
```{r}
# Outputs the min and max
range(who$life_expectancy)
```

```{r}
max(who$life_expectancy) - min(who$life_expectancy)

```

National average life expectancies in 2015 **ranged from 51 to 88**, for a total difference of **37 years**.



---
##### Fodder

---
# Central Tendency and Asymmetry 

* <span style="color: red;">Mean</span>
* <span style="color: blue;">Median</span>
* <span style="color: purple;">Mode</span>

```{r, echo = F}
pos_vlines <- pos_skew + 
  geom_vline(xintercept = mean(x$pos_skew), color = "red") +
  geom_vline(xintercept = median(x$pos_skew), color = "blue") +
  geom_vline(xintercept = getmode(x$pos_skew), color = "purple")

neg_vlines <- neg_skew + 
  geom_vline(xintercept = mean(x$neg_skew), color = "red") +
  geom_vline(xintercept = median(x$neg_skew), color = "blue") +
  geom_vline(xintercept = getmode(x$neg_skew), color = "purple")

normal_vlines <- normal +
  geom_vline(xintercept = mean(x$central), color = "red") +
  geom_vline(xintercept = median(x$central), color = "blue") +
  geom_vline(xintercept = getmode(x$central), color = "purple")

gridExtra::grid.arrange(pos_vlines, normal_vlines, neg_vlines, nrow = 1)

```

---
# Variability
## Variance and Standard Deviation

* Because variance and standard deviations are calculated from the mean, they are sensitive to skew and outliers (just like the mean).

WORTH ADDING VISUALS? (PROBABLY NOT THESE)
```{r}
ggplot(x, aes(x = pos_skew)) +
  geom_density(adjust = 3) +
  geom_vline(xintercept = mean(x$pos_skew), color = "red", linetype = "dotted") +
  geom_vline(xintercept = mean(x$pos_skew) - sd(x$pos_skew), color = "cornflowerblue") +
  geom_vline(xintercept = mean(x$pos_skew) + sd(x$pos_skew), color = "cornflowerblue")

ggplot(x, aes(x = neg_skew)) +
  geom_density(adjust = 3) +
  geom_vline(xintercept = mean(x$neg_skew), color = "red", linetype = "dotted") +
  geom_vline(xintercept = mean(x$neg_skew) - sd(x$neg_skew), color = "cornflowerblue") +
  geom_vline(xintercept = mean(x$neg_skew) + sd(x$neg_skew), color = "cornflowerblue")

ggplot(x, aes(x = central)) +
  geom_density(adjust = 3) +
  geom_vline(xintercept = mean(x$central), color = "red", linetype = "dotted") +
  geom_vline(xintercept = mean(x$central) - sd(x$central), color = "cornflowerblue") +
  geom_vline(xintercept = mean(x$central) + sd(x$central), color = "cornflowerblue")

```
---
## Visualizing Variance
ADD CONTENT ON BOXPLOTS, ETC.

---
```{r}
map(x, ~ ggplot(x, aes(.x)) +
  geom_density() +
  theme_minimal() +
  geom_vline(xintercept = mean(.x), color = "red") +
  geom_vline(xintercept = median(.x), color = "green", type = "dash") +
  geom_vline(xintercept = getmode(.x), color = "blue"))

ggplot(x, aes(pos_skew)) +
  geom_density() +
  geom_vline(xintercept = mean(pos_skew))

ggplot(x, aes(neg_skew)) +
  geom_density()

```