---
title: "EDUC 641 Continuous Data"
author: "Professor Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
  html_document:
    keep_md: true
---
class: inverse, center, middle

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.align = "center",
                      fig.height = 3)

if (!require(pacman)) install.packages('pacman', repos = 'https://cran.rstudio.com')
pacman::p_load(tidyverse, here, knitr, dplyr, ggplot2, blogdown, rio)
needs(gifski)
```

```{r, echo = F}
who <- import(here::here("data", "life_expectancy.csv")) %>% 
  janitor::clean_names() %>% 
  filter(year == 2015) %>%
  select(country, status, year, life_expectancy) %>% 
  rename(region = country) %>% 
  mutate(life_expectancy = round(life_expectancy, digits = 0))

```

# Agenda
* Sampling Distributions
* Central Limit Theorem

---
# Review
## Distributions

Characterized by:

* Central Tendency
+ Mean
+ Median
+ Mode

* Variability
+ Range
+ Interquartile Range
+ Variance
+ Standard Deviation

* Shape
+ Skew
+ Kurtosis

---
# Review
## Normal Distributions

A normal distribution is a **theoretical** distribution with the following characteristics:

Central Tendency:
+ The mean, median, and mode are equivalent values.

Variability:
+ 68% of observations are within 1 SD
+ 95% of  observations are within 2 SDs
+ 99% of observations are within 3 SDs

Shape: 
+ Symmetrical

---
# Review
## Normal Distribution

* The area under a normal distribution is always equal to 1.
* The area under the curve corresponds to the proportion of observations.

```{r}

ggplot(aes, )
```

# Population vs. Sample Distributions
 
Using the WHO dataset, let's say we want to identify the mean life expectancy across all countries. However, let's also assume we only have the resources to sample from 20 countries.

.pull-left[

Our **population statistic** represents the true value across all countries.

```{r, echo = F}
world <- map_data("world")

regions <- world %>% 
  distinct(region) %>% 
  sample_n(10) %>% 
  mutate(sample = 1)

world <- left_join(world, regions) %>% 
  mutate(population = 1)
```

```{r, echo = F, fig.height=3, fig.width= 4}
ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region, fill = as.factor(population)),
    color = "black", size = 0.1
  ) +
  theme(legend.position = "none")
```
]

.pull-right[
Our **sample statistic** will be our estimate of the population statistic using the countries we examine.

```{r, echo = F, fig.height = 3, fig.width = 4}
ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region, fill = as.factor(sample)),
    color = "black", size = 0.1
  ) +
  theme(legend.position = "none")
```
]

---
# Sample Estimates

Assume that our true population statistic is 71.64. 

```{r}
mean(who$life_expectancy)
```

```{r, echo = F}
sim_num <- seq(1:100)
region_samp <- vector("list", 100)
for (i in seq_along(region_samp)){
    region_samp[[i]] <- who %>% 
      filter(str_length(region) <= 12) %>% 
      sample_n(10)
}

names(region_samp) <- paste0("sim_", sim_num)
```

```{r, echo = F, fig.width = 8, fig.height = 5}
plots <- lapply(region_samp[1:4], function(x) {
    ggplot(x, aes(x = region, y = life_expectancy)) +
    geom_point() +
    geom_hline(yintercept = mean(x$life_expectancy), color = "red", linetype = "dotted") +
    geom_hline(yintercept = mean(who$life_expectancy), color = "blue") +
    geom_label(aes(x = +Inf, y = mean(x$life_expectancy), vjust = 2, label =  paste(mean(x$life_expectancy))), colour="red", size = 4) +
        coord_flip() +
    theme_minimal()
})
library(gridExtra)
marrangeGrob(plots, nrow=2, ncol=2)
```

---


---
# Population vs. Sample Distribution

```{r}
samp_means <- tibble(samp = 1:100, means = map_dbl(region_samp, ~mean(.$life_expectancy)))
```


```{r, animation.hook="gifski", interval = 0.4, echo = F}
for (i in 1:100) {
  x <- ggplot(subset(samp_means, samp <= i), aes(means)) +
           geom_histogram(fill = "red") +
           geom_vline(xintercept = 71.64, color = "blue") +
  xlim(65, 85) +
  xlab("Mean Life Expectancy") +
  theme_minimal()
  print(x)
}
```
