---
title: "EDUC 641 Continuous Data"
author: "Professor Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
  html_document:
    keep_md: true
---
class: inverse, center, middle

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.align = "center",
                      fig.height = 3)

if (!require(pacman)) install.packages('pacman', repos = 'https://cran.rstudio.com')
pacman::p_load(tidyverse, here, knitr, dplyr, ggplot2, blogdown, rio)


```

```{r, echo = F}
who <- import(here::here("data", "life_expectancy.csv")) %>% 
  janitor::clean_names() %>% 
  filter(year == 2015) %>% 
  select(country, status, life_expectancy) %>% 
  rename(region = country) %>% 
  mutate(life_expectancy = round(life_expectancy, digits = 0))
world <- map_data("world")

who1<- unique(who$region)
world1 <- unique(world$region)


```

# Agenda
* Objectives
* Life Expectancy Data
* Measures of Central Tendency
* Measures of Variability


---
# Life Expectancy Data
Suppose you are working for the World Health Organization and are investigating life expectancy across different regions. 

.bg-washed-green.b--dark-green.ba.bw2[
Using this dataset, we can ask questions like:
* How does life expectancy compare in developing vs. developed regions?
* What is the relationship between average years of schooling and life expectancy?
]

**But**, before making comparisons between social/economic conditions, we want to start with describing the full dataset, looking at all 183 world regions.

**Our Task:** Describe the distribution of regional life expectancy in 2015.

---
## Start with visualized versions and then proceed to mathematical examples.
## Potentially give examples where mean and median differnences matter/don't matter
# Describing a Distribution

There are many ways to describe the characteristics of a distribution. 

Today we are reviewing:

Measures of Central Tendency
  * Mean
  * Median
  * Mode
  
Spread or Variability
  * Variance
  * Standard Deviation
  * Interquartile Range
  * Range
  
---
# But first, summation notation

Summation (or sigma) notation is used to provide a concise expression for the sum of observations.

$$\sum_{i=i}^{n} x_{i}$$

---
# Central Tendency
## Mean
Represents the average, or the sum of all observations divided by the number of observations
 *$$\mu = \frac{\sum_{i=i}^{n} x_{i}} {n}$$
 $$\sum_{n=1}^{10} n^2$$
 

```{r}
mean(who$life_expectancy)

```

The average national life expectancy in 2015 was **71.62** years.

```{r, echo = F}

sim_data <- data_frame(life_expectancy = rnorm(n = 100000, mean = mean(who$life_expectancy), sd = sd(who$life_expectancy)))

ggplot(sim_data, aes(x = life_expectancy)) +
  geom_density() +
  geom_vline(xintercept = 71.62, color = "red") +
  theme_minimal()

ggplot(who, aes(x = life_expectancy)) +
  geom_histogram() +
  geom_vline(xintercept = 71.62, color = "red") +
  theme_minimal()


```

---
## Median
* The median represents the midpoint of the distribution.
* It is also called the 50th percentile of a distribution, meaning half of the observations are above the median and half are below.
* The median is particularly useful for describing the central tendency of skewed distributions (more on this later).

```{r}
median(who$life_expectancy)
```

The median national life expectancy in 2015 was **73.9** years.
---
## Mode (skip over how to do it in R?)
The mode refers to the most frequent value in the data. 
```{r, echo = F}
ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "black", fill = "lightgray", size = 0.3
  ) +
  theme_minimal()
samp1 <- who %>% 
  sample_frac(.10) %>% 
  mutate(samp1 = 1)

world <- left_join(world, who)

worldplot <- ggplot(data = world, aes(x=long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(aes(fill = status))
worldplot

mode(who)
```

---
# Variability
* Variance
* Standard Deviation
* Interquartile Range
* Range

---
# Variance

 *$$\(s^{2} = \frac{\sum_{i=i}^{n} x_{i}}{n} \)$$

```{r}
world <- world %>% 
  distinct(region)


```

